Gaia's constraints on the extremely massive ($\sim$$150\,M_\odot$) runaway star candidate VFTS682

### 2018-06-13 VFTS682_APJL.tex
minor

    \end{tabular}
    \tablecomments
    { 
      (1)~\cite{schneider:18}
      (2)~\cite{bestenlehner:11}
      (3)~\cite{rubio-diez:17}
    }
  \end{center}
  \label{tab:star}
\end{table}

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
%
%of spectral type WNh5 using multi-epoch spectra covering $\lambda$4000--7000 from the
%VFTS survey. % The spectra available covered $\lambda$4000--7000 and
% were taken at multiple epochs.
%
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of the brightest stars in the region. The star is unlikely to have a close companion, because of the absence of clear radial velocity variations \citet{bestenlehner:11}.  Bayesian fits of the stellar parameters against evolutionary tracks \citep{brott:11, kohler:15} using the Bonsaii code \citep{schneider:17, schneider:18} provide estimates for the age, present mass and initial mass, indicating that this star among the most massive stars known, see \Tabref{tab:star} for an overview of the parameters.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.\todo{possibly: cartoon to illustrate angle definition, or add to
    fig.2 if not too messy}}
  \label{fig:dist}
\end{figure*}

@@

The left panel shows the distribution in two-dimensional proper motion relative
to R136, and shows that VFTS682 is not a peculiar star in terms of
its motion relative to R136. Subtracting the mean motion of the field,
we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both indicate
a non-zero peculiar projected motion that would make it a slow runaway
star as expected by \cite{bestenlehner:11} if the star was ejected
from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
More importantly, both the proper motion
measurement from \emph{Gaia} and HST fall in the range expected to
reach its present day location % at $\sim$29\,pc from the cluster core
if the
star was ejected very early in its evolution. This range is
highlighted by the green vertical band, and its width is determined by
the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star}).

@@

Figure~\ref{fig:main} shows the projected motion of VFTS682 relative to R136 on the
sky. \todo{fix colors} The dashed circle shows the location of the cluster, and the red
(cyan) arrow shows the relative proper motion from \emph{Gaia} DR2
(HST). The lighter arrows show the corresponding opening angle within
the large uncertainties. The length of each arrow is proportional to
the speed of the star, and dashed prolungations show the possible
region of origin assuming an age of $1.0\pm0.2$\,Myr
\citep[][]{schneider:18}. This figure illustrates that R136 is clearly
the most likely origin of the star, although the large error bars
prevent a robust identification.

Assuming VFTS682 originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

@@

Based on our results, we tentatively claim that VFTS682 is the most massive
(slow) runaway known to date, with a peculiar three-dimensional speed of
$44\pm21\,\kms$ \todo{make 2d? use LOS velocity from Marimar?}. Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.
If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

@@

\section*{Affiliations}
\noindent $^{1}${Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands} \\
  $^{2}$ {ESA, European Space Astronomy Centre, Apdo. de Correos 78,
    E-28691 Villanueva de la Ca\~nada, Madrid, Spain} \\
  $^{3}$ {Space Telescope Science Institute, 3700 San Martin Drive,
    Baltimore, MD 21218, USA}\\
  $^{4}$ {Center for Astrophysical Sciences, Department of Physics \& Astronomy, Johns Hopkins University, Baltimore, MD 21218, USA}\\
  $^{5}${Department of Physics and Astronomy, Hicks Building,
    Hounsfield Road, University of Sheffield, Sheffield S3 7RH, UK}\\
  $^{6}${UK Astronomy Technology Centre, Royal Observatory Edinburgh, Blackford Hill, Edinburgh, EH9 3HJ, UK}\\
  $^{7}$ {National Research Council, Herzberg Astronomy \&
    Astrophysics, 5071 West Saanich Road, Victoria, BC, V9E 2E7,
    Canada}\\
  $^{8}$ {School of Astronomy \& Space Science, University of the Chinese
    Academy of Sciences, Beijing 100012, China}\\
  $^{9}$ {National Astronomical Observatories, Chinese Academy of
    Sciences, Beijing 100012, China}\\
  $^{10}$ {Argelander-Instit\"ut f\"ur Astronomie, Universit\"at Bonn,
    Auf dem H\"ugel 71, 53121, Bonn, Germany}\\
  $^{11}$ {Centro de Astrobiología, CSIC-INTA, Carretera de Torrejón a Ajalvir km-4, E-28850 Torrejón de Ardoz, Madrid, Spain}\\
  $^{12}$ {Institute of Astronomy, KU Leuven, Celestijnenlaan 200 D, B-3001 Leuven, Belgium}\\
  $^{13}$ {Department of Physics, University of Oxford, Keble Road,
    Oxford OX1 3RH, UK} \\
  $^{14}$ {Armagh Observatory, College Hill, Armagh BT61 9DG, UK}\\
}

>>>>

    \end{tabular}
    \tablecomments
    { 
      (1)~\cite{schneider:18}
      (2)~\cite{bestenlehner:11}
      (3)~\cite{rubio-diez:17}
    }
  \end{center}
  \label{tab:star_param}
\end{table}

@@

The WNh5 star VFTS682, located at right ascension (RA)
05$^\mathrm{h}$38$^\mathrm{m}$55.510$^\mathrm{s}$  and declination
(DEC) \mbox{-69$^\mathrm{o}$04'26.72''} J2000, was observed as part of the multi-epoch, spectroscopic VFTS campaign covering $\lambda$4000--7000 \citep[][]{evans:11}. 
%
%of spectral type WNh5 using multi-epoch spectra covering $\lambda$4000--7000 from the
%VFTS survey. % The spectra available covered $\lambda$4000--7000 and
% were taken at multiple epochs.
%
\citet{bestenlehner:11}  analyzed the spectra to infer the stellar parameters and measure an extinction of $A_V=4.45\pm0.12$, implying a luminosity of $\log_{10}(L/L_\odot) =  6.5\pm0.2$, making this one of the brightest stars in the region. The star is unlikely to have a close companion, because of the absence of clear radial velocity variations \citet{bestenlehner:11}.  Bayesian fits of the stellar parameters against evolutionary tracks \citep{brott:11, kohler:15} using the Bonsaii code \citep{schneider:17, schneider:18} provide estimates for the age, present mass and initial mass, indicating that this star among the most massive stars known, see \Tabref{tab:star_param} for an overview of the parameters.

@@

\begin{figure*}[htbp]
  \centering
  \includegraphics[width=0.35\textwidth]{figures/dist_mu_region.pdf}
  \includegraphics[width=0.35\textwidth]{figures/angle}
  \caption{\emph{Left panel}: distribution of OB-type and Wolf-Rayet stars in proper
    motion relative to R136. VFTS682 is not an outlier, but
    its relative proper motion matches the expected value if it were indeed
    ejected from R136 assuming an age of $1.0\pm0.2$\,Myr. The top axis shows the conversion to physical units
    assuming a distance of 50\,kpc. \emph{Right panel}:  distribution of
    angles between the relative proper motion direction and the radial
    direction from the center of the cluster to the star. The error bars from \emph{Gaia} for VFTS682 are large, but
    the best value is in agreement with the hypothesis of dynamical
    ejection. In both
    panels, the dark blue histograms contain 317 
    stars with error smaller than $0.1\,\mathrm{mas \
      yr^{-1}}\simeq25\,\mathrm{km\ s^{-1}}$ at 50\,kpc, the
    lighter blue histograms contain 36 stars with errors smaller than $0.05\,\mathrm{mas \
      yr^{-1}}$.\todo{possibly: cartoon to illustrate angle definition, or add to
    fig.2 if not too messy, maybe also show HST measurements on this
    plot for 682?}}
  \label{fig:dist}
\end{figure*}

@@

The left panel shows the distribution in two-dimensional proper motion relative
to R136, and shows that VFTS682 is not a peculiar star in terms of
its motion relative to R136. Subtracting the mean motion of the field,
we obtain $\delta \mu_{Gaia}=0.13\pm 0.09\masyr (32\pm 21\kms)$ and
$\delta \mu_{HST}=0.20\pm 0.10\masyr (47 \pm 24\kms )$. Both indicate
a non-zero peculiar projected motion that would make it a slow runaway
star as expected by \cite{bestenlehner:11} if the star was ejected
from the cluster,  but in both cases the meaurements are also consistent with zero within 2$\sigma$. 
More importantly, both the proper motion
measurement from \emph{Gaia} and HST fall in the range expected to
reach its present day location % at $\sim$29\,pc from the cluster core
if the
star was ejected very early in its evolution. This range is
highlighted by the green vertical band, and its width is determined by
the uncertainty in the present-day age of the star (cf.~\Tabref{tab:star_param}).

@@

Figure~\ref{fig:main} shows the projected motion of VFTS682 relative to R136 on the
sky. \todo{fix colors} The dashed circle shows the location of the cluster, and the red
(cyan) arrow shows the relative proper motion from \emph{Gaia} DR2
(HST). The lighter arrows show the uncertainty on the direction of
motion due to the large uncertainties. The length of each arrow is proportional to
the speed of the star, and dashed prolungations show the possible
region of origin assuming an age of $1.0\pm0.2$\,Myr
\citep[][]{schneider:18}. This figure illustrates that R136 is clearly
the most likely origin of the star, although the large error bars
prevent a robust identification.

Assuming VFTS682 indeed originates from R136, we can calculate its kinematic
age simply as:
\begin{equation}
  \label{eq:kin_age}
  \tau_\mathrm{kin} = \frac{d_\parallel}{\delta\mu_{Gaia}} \simeq
  \frac{119.4\,\mathrm{arcsec}}{0.13\masyr} \simeq 0.9\pm\,0.6\, \mathrm{Myr} \ \ ,
\end{equation}
where $d_\parallel = 119.4\,\mathrm{arcsec}$ is the angular distance from VFTS682 to
the core of the cluster \citep[corresponding to $\sim$29\,pc at LMC distance,][]{bestenlehner:11}.
As in the rest of this study, we neglect for
simplicity the error on the distance estimates, because it is negligible compared to other uncertainties.
The kinematic age $\tau_\mathrm{kin}$ is compatible with a very early
ejection from the cluster, given is apparent age.
% of $1.0\pm 0.2$\,Myr 
% \citep{schneider:18}.

@@

Based on our results, we tentatively claim that VFTS682 is the most massive
(slow) runaway known to date, with a peculiar two-dimensional
projected velocity with respect to R136 of
$\sim38\pm22\kms$ (averaging the \emph{Gaia} DR2 and HST
results\todo{check numbers}). Due to the large error bars, this result will need
to be revisited with future astrometric data. % to be revisited as updated
% astrometric parameters from future \emph{Gaia} data releases become
% available.
If confirmed, it means that isolated star formation is
\emph{not} required to explain the isolation of VFTS682. Its proper motion suggests that it was ejected from the cluster R136
$0.9\pm0.6$\,Myr ago. Because of the exceptionally large mass
of this star, this raises the question of which stars must populate
the core of the cluster.

@@

\section*{Affiliations}
\noindent $^{1}${Astronomical Institute Anton Pannekoek, University of
    Amsterdam, 1098 XH Amsterdam, The Netherlands} \\
  $^{2}$ {ESA, European Space Astronomy Centre, Apdo. de Correos 78,
    E-28691 Villanueva de la Ca\~nada, Madrid, Spain} \\
  $^{3}$ {Space Telescope Science Institute, 3700 San Martin Drive,
    Baltimore, MD 21218, USA}\\
  $^{4}$ {Center for Astrophysical Sciences, Department of Physics \& Astronomy, Johns Hopkins University, Baltimore, MD 21218, USA}\\
  $^{5}${Department of Physics and Astronomy, Hicks Building,
    Hounsfield Road, University of Sheffield, Sheffield S3 7RH, UK}\\
  $^{6}${UK Astronomy Technology Centre, Royal Observatory Edinburgh, Blackford Hill, Edinburgh, EH9 3HJ, UK}\\
  $^{7}$ {National Research Council, Herzberg Astronomy \&
    Astrophysics, 5071 West Saanich Road, Victoria, BC, V9E 2E7,
    Canada}\\
  $^{8}$ {School of Astronomy \& Space Science, University of the Chinese
    Academy of Sciences, Beijing 100012, China}\\
  $^{9}$ {National Astronomical Observatories, Chinese Academy of
    Sciences, Beijing 100012, China}\\
  $^{10}$ {Argelander-Instit\"ut f\"ur Astronomie, Universit\"at Bonn,
    Auf dem H\"ugel 71, 53121, Bonn, Germany}\\
  $^{11}$ {Centro de Astrobiología, CSIC-INTA, Carretera de Torrejón a Ajalvir km-4, E-28850 Torrejón de Ardoz, Madrid, Spain}\\
  $^{12}$ {Institute of Astronomy, KU Leuven, Celestijnenlaan 200 D, B-3001 Leuven, Belgium}\\
  $^{13}$ {Department of Physics, University of Oxford, Keble Road,
    Oxford OX1 3RH, UK} \\
  $^{14}$ {Armagh Observatory, College Hill, Armagh BT61 9DG, UK}\\
